$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-04-10 write_up/ms.tex
edited abstract and sections 1-2

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2018}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}
\usepackage{algorithm,algorithmic}
\usepackage{graphicx}
\usepackage{bbm}

@@

\begin{abstract}
	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Despite its importance, MCMC is a notoriously subtle beast. Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models. Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a stationary distribution. Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred. Here, we introduce a new method for probing convergence based on training machine learning algorithms to classify samples according to the chain that generated them: we call this convergence measure $R^*$. In contrast to the predominant $\hat{R}$, $R^*$ is a single statistic across all parameters that captures whether convergence has occurred, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distributions, which is currently largely overlooked by existing approaches. Since our machine learning method, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
\end{abstract}

\section{Introduction}
Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years. In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc} and the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, which Stan \citep{carpenter2017stan} and PyMC3 \citep{salvatier2016probabilistic} implement. MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.

Its importance in applied scientists' toolkits means it is essential that MCMC is used properly and with adequate care. A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior. Because of this, software packages (Stan in particular \citep{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling. Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use. We repeat here a quote from Chris Holmes to this effect \citep{holmes},

\begin{quote}
	MCMC is most definitely not a panacea and should be used with caution, almost as a last resort. It is just that often we are at the last resort for interesting models.
\end{quote}

The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{lambert2018Student}. MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions). With only these asymptotic guarantees, it is actually quite miraculous that MCMC samplers \textit{can} often approximate the posterior with only a (relatively small) finite number of samples.

@@

To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \citep{gelman1992inference}. If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge. This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \citep{lambert2018Student}. The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\hat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \citep{gelman2013bayesian}. If these variances are similar, $\hat{R}\approx 1$, and chains are deemed to have mixed. Recently, Stan has adopted more advanced variations on the original $\hat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{carpenter2017stan}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\hat{R}$ \citep{vehtari2019rank}. Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \citep{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}. Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample's value the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a sample. In this vein, we use a supervised machine learning (ML) approach to measure convergence. Specifically, we train a ML model to classify the chain that generated each observation. By evaluating the performance of the model on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, ML models naturally exploit differences between the full joint distributions of each chain, which means they are sensitive to variations across the joint distribution of target model dimensions unlike most existent approaches. This statistic, unlike its $\hat{R}$ cousins, is univariate: one model provides a single $R^*$, whereas $\hat{R}$ has separate values for each parameter. However, the ML models we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our machine learning model, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\hat{R}$ measures calculated (typically less than 1 second to calculate). It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

@@

These two cases, whilst simple, illustrate the basis of our approach. To determine if a set of Markov chains has converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs. By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains. By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \citep{gelman2013bayesian}. In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence. Algorithm \ref{alg:R_star} gives a recipe for calculating $R^*$.

@@

The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing. Our GBM model was rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters. In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

A benefit of most ML approaches to classification is that they can output class probabilities, opposed to a single predicted class, which we leverage to produce an uncertainty distribution for $R^*$. Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating samples from this distribution, which we now elaborate on in words. For each sample, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$. By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct. We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$. We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.

@@

First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\hat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS. In all cases, the various quantities were calculated based on chains split into halves. For both dimensions, the two ESS measures were above 3700 and $\hat{R}<1.001$, indicating no issues with convergence.

@@

We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5). First, we used the ``centered'' parameterisation of this model, which is of the form,
%
\begin{equation}\label{eq:mvt_normal_250}
\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
\end{equation}
%
where $\boldsymbol{x}\in\mathbb{R}^{250}$. For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.02): in this case, $\hat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.

Rather than run the MCMC sampler for more iterations, we move to a ``non-centered'' parameterisation, which introduces auxillary variables $\boldsymbol{z}\in\mathbb{R}^{250}$ that don't affect $p(\boldsymbol{x})$ but facilitate sampling from it. This model has the form,
%
\begin{align}
\boldsymbol{A}^{-1} = \boldsymbol{L}\boldsymbol{L}^T,\qquad
\boldsymbol{x} = \boldsymbol{L} \boldsymbol{z},\qquad
z_j\sim \mathcal{N}(0, 1), \text{ for } j = 1,2,...,250.
\end{align}
%
where $\boldsymbol{L}$ is the Cholesky decomposition of the covariance matrix, $\boldsymbol{A}^{-1}$. Fig. \ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is centred on $R^*=1$ (mean $R^*=1.00$) indicating convergence. This was also reiterated by the fact that $\hat{R}<1.01$ and $\text{ESS}>400$ for all parameters.

In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power. We now compare these with the more established metrics $\hat{R}$ and ESS as calculated in Stan. For a GBM fitted to the centered model, we plot in Fig. \ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\hat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle). From this plot, there appears a weak positive association between GBM's variable importance and $\hat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$). In Fig.  \ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$). Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\hat{R}$ and ESS.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\hat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan. In plots B and C, horizontal jitter was added to the points and a loess fit line with standard errors overlaid. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:mvt}
\end{figure}

\subsection{Wide datasets: multivariate normal}\label{sec:wide}
As the number of parameter dimensions increases, it could be hypthesised that ML algorithms will overfit the data, and, hence, testing set classification would be poor; in being so, this could lead us to falsely conclude that convergence had occurred. To test this hypothesis, we investigated two scenarios based around the multivariate normal. In the first of these, we used the 250-dimensional multivariate normal of eq. \ref{eq:mvt_normal_250} and used 250 post-warm-up iterations (after 250 warm-up iterations) from Stan to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}. Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig. \ref{fig:mvt_wide_both}A. In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence. The same conclusion is reached if split-$\hat{R}$ is used instead (Fig. \ref{fig:wide_both_diagnostics}A), since, for both parameterisations, there remain some parameters with $\hat{R}>1.01$. Using bulk- or tail-ESS instead, we conclude that the non-centered parameterisation shows signs of convergence whereas the centered does not (Fig. \ref{fig:wide_both_diagnostics}B\&C).

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{../output/wide_both_diagnostics.pdf}}
	\caption{\textbf{Established convergence diagnostics for 250-dimensional normal example.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered. Column A shows split-$\hat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up samples from Stan's NUTS algorithm. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:wide_both_diagnostics}
\end{figure}

We next consider a more challenging example -- a target distribution with 10,000 dimensions. In this case, we assume independent standard normals for each dimension. In Fig. \ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distributions: one with 400 iterations, the other with 1000. In both cases, the distributions were right of $R^*=1$, indicating non-convergence. These results were also echoed by split-$\hat{R}$, with 28\% of dimensions having $\hat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.

>>>>

% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{neurips_2018}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true,
            colorlinks=true,
            linkcolor=blue,
            citecolor=blue,
            filecolor=blue,
            urlcolor=blue}
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}
\usepackage{algorithm,algorithmic}
\usepackage{graphicx}
\usepackage{bbm}

\newcommand{\aki}[1]{\textcolor{red}{[Aki: #1]}}

@@

\begin{abstract}
  Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists.
% Despite its importance, MCMC is a notoriously subtle beast. Indeed, the words of statisticians often echo that MCMC ``should be used with caution, almost as a last resort''; the problem is that we are usually at the ``last resort'' for interesting models. Central to these concerns is the difficulty in determining whether Markov chains have converged to the posterior distribution.
  Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about its finite sample performance. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution.
%  Qualitatively, these summary statistics aim to determine whether it is possible to predict the chain that generated a particular sample: if these predictions are accurate, then the chains have not mixed and convergence has not occurred.
Here, we introduce a new method for diagnosing convergence based whether machine learning classifier model can successfully discriminate the individual chains. We call this convergence measure $R^*$. In contrast to the predominant $\widehat{R}$, $R^*$ is a single statistic across all parameters that indicates lack of mixing, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; instead using all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches. Since the used machine learning classifier, gradient-boosted regression trees (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
\end{abstract}

\section{Introduction}
Markov chain Monte Carlo (MCMC) is the class of exact-approximate methods that has contributed most to applied Bayesian inference in recent years. In particular, MCMC has made Bayesian inference widely available to a diverse community of practitioners through the many software packages that use it as an internal inference engine: from Gibbs sampling \citep{geman1984stochastic}, which underpins the popular BUGS \citep{lunn2000winbugs} and JAGS \citep{plummer2003jags} libraries, to more recent algorithms: for example, Hamiltonian Monte Carlo (HMC) \citep{neal2011mcmc}, the No U-Turn Sampler (NUTS) \citep{hoffman2014no}, and a dynamic HMC variant \citep{betancourt2017conceptual}, which Stan \citep{carpenter2017stan} and PyMC3 \citep{salvatier2016probabilistic} implement. MCMC methods are currently the most effective tools for sampling from many classes of posterior distributions encountered in applied work, and it seems unlikely that this trend will change soon.

Its importance in applied scientists' toolkits means it is essential that MCMC is used properly and with adequate care. A cost of automated inference software is that it is increasingly easy to regard MCMC as oracular: giving uncompromised views onto the posterior. Because of this, software packages (Stan in particular \citep{carpenter2017stan} is a great exemplar of this), go to great lengths to communicate to users any issues with sampling.
% Indeed, in spite of this ease-of-application, many statisticians who develop MCMC methods are less at ease with recommending their use. We repeat here a quote from Chris Holmes to this effect \citep{holmes},
% \begin{quote}
% 	MCMC is most definitely not a panacea and should be used with caution, almost as a last resort. It is just that often we are at the last resort for interesting models.
% \end{quote}

The most important determination of whether MCMC has worked is whether the sampling distribution has converged to the posterior \citep{lambert2018Student}.\aki{Maybe some other reference, too?} MCMC methods are thus created because of an asymptotic property: that given an infinite number of samples, their sampling distribution approaches the posterior (under general conditions). Although the guarantees are asymptotic, MCMC estimates can have negligible bias with only a relatively small finite number of samples.

@@

To combat this curse of hindsight, running multiple, independent chains, which have been initialised at diverse areas of parameter space is recommended \citep{gelman1992inference}. If the chains appear not to ``mix'' -- a term essentially meaning that it is difficult to resolve an individual chain's path from the mass of paths overlaid on top of one another -- they are yet to converge. This approach makes it less likely that faux-convergence will occur due to chains becoming stuck in an area of parameter space, and running multiple chains is standard practice in applied inference \citep{lambert2018Student}. The predominant approach to quantitatively measuring this mixing is to compare each chain's sampling distribution to that of the population of chains as a whole: specifically, $\widehat{R}$ -- the main convergence statistic used -- compares within-chain variance to that between-chains \citep{gelman1992inference}. If these variances are similar, $\widehat{R}\approx 1$, and chains are deemed to have mixed. Recently, Stan has adopted more advanced variations on the original $\widehat{R}$ formula: for example, splitting individual chains in two to combat poor intra-chain mixing \citep{gelman2013bayesian}; and using ranks of parameter samples rather than the raw sample values themselves to calculate $\widehat{R}$ \citep{vehtari2019rank}. Additionally, there has been more focus on ensuring that the effective sample size (ESS), a measure of sample quality (see, for example, \citep{lambert2018Student}), is sufficient, and accordingly, new measures of this quantity have been proposed \citep{vehtari2019rank} and adopted \citep{carpenter2017stan}. Collectively, these statistics help alert users of MCMC to issues with sampling (that typically echo issues with the model) meaning that all is not hunky dory.

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a sample's value the chain that generated it. Rephrased, it should not be possible to predict the chain that \textit{caused} a sample. In this vein, we use a machine learning (ML) classifier to measure convergence. Specifically, we train a ML model to predict the chain that generated each observation. By evaluating the performance of the model on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, the used ML classifier naturally exploit differences between the full joint distributions of each chain, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics. This statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution. However, the ML classifier we use can straightforwardly be interrogated to determine which parameters were most important for generating predictive accuracy. For our machine learning classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically less than 1 second to calculate). It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

@@

These two cases, whilst simple, illustrate the basis of our approach. To determine if a set of Markov chains has converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs. By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains. By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\widehat{R}$ \citep{gelman2013bayesian}. In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence. Algorithm \ref{alg:R_star} gives a recipe for calculating $R^*$.

@@

The ML model we use here is a gradient-boosted regression tree (also known as a type of gradient-boosted machine or GBM, introduced in \cite{friedman2001greedy}), which experience has dictated to be a highly predictive framework for use in tabular data \citep{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s ``Caret'' package \citep{kuhn2008building}, which, in turn, uses the ``gbm'' package \citep{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{S}\times \mathbb{R}^{K}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's samples into randomly divided training and testing tests: here, we use 70\% of samples for training and 30\% for testing. Our GBM model was rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for both these steps for most models we consider in \S\ref{sec:results}), and its predictive performance was insensitive to its hyperparameters. In all the examples explored in \S\ref{sec:results}, the GBM hyperparameter settings we used were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations being the minimum required for each node, and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

We can also get class probabilities, which we leverage to produce an uncertainty distribution for $R^*$. Algorithm \ref{alg:R_star_uncertainty} gives a recipe for generating samples from this distribution, which we now elaborate on in words. For each sample, $s$, in our testing set, GBM outputs a simplex of chain probabilities: $\boldsymbol{p}^{\{s\}}=(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$, which forms a categorical distribution that can be sampled from to yield a unique chain prediction, $c^{\{s\}}$. By comparing this classification to the true classification, $c^s$, we obtain a binary measure, $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$, of whether this prediction was correct. We repeat this process for each sample in the testing set, generating $\boldsymbol{a}=(a^{\{1\}},a^{\{2\}},...,a^{\{N_\text{test}\}})$, whose average yields a single $R^{*\{i\}}=N \bar{a}$ estimate for iteration $i$. We then iterate this process, for $i=1,2,...,I$, producing a set of $(R^{*\{1\}},R^{*\{2\}},...,R^{*\{I\}})$, which collectively represent a distribution for $R^*$.

@@

First, we use the code provided in \cite{vehtari2019rank} to calculate rank-normalised $\widehat{R}$ and two different ESS measures, aiming to capture how well certain regions of the posterior have been explored: these are known as bulk ESS and tail ESS. In all cases, the various quantities were calculated based on chains split into halves. For both dimensions, the two ESS measures were above 3700 and $\widehat{R}<1.001$, indicating no issues with convergence.

@@

We next consider a more challenging problem -- a 250-dimensional multivariate normal target where its precision matrix, $\boldsymbol{A}\in\mathbb{R}^{250}\times\mathbb{R}^{250}$, is generated from a Wishart distribution \citep{hoffman2014no}. We assume that the Wishart distribution's degrees of freedom parameter is 250, resulting in a distribution with high correlations between dimensions. We use Stan's NUTS algorithm to sample from this target distribution, and run the algorithm for two different iteration counts (each time across 4 chains): 400 and 10,000 (the latter thinned by a factor of 5). First, we used the ``centered'' parameterisation of this model, which is of the form,
%
\begin{equation}\label{eq:mvt_normal_250}
\boldsymbol{x}\sim \mathcal{N}(\boldsymbol{0},\boldsymbol{A}^{-1}),
\end{equation}
%
where $\boldsymbol{x}\in\mathbb{R}^{250}$. For each set of samples, we used Algorithm \ref{alg:R_star_uncertainty} to generate an uncertainty distribution for $R^*$, which are shown in Fig. \ref{fig:mvt}A. From the plot for the 400 iteration case, it is clear that convergence has not yet occurred since $R^*>1$ across the bulk of this distribution. Even in the 10,000 iteration case, the $R^*$ distribution remains stubbornly shifted a little rightwards of $R^*=1$ (its mean is 1.02): in this case, $\widehat{R}<1.01$ for all parameters, although 61\% of parameters had an $\text{ESS}<400$ indicating issues with convergence \citep{vehtari2019rank}.

Rather than run the MCMC sampler for more iterations, we move to a ``non-centered'' parameterisation, which introduces auxillary variables $\boldsymbol{z}\in\mathbb{R}^{250}$ that don't affect $p(\boldsymbol{x})$ but facilitate sampling from it. This model has the form,
%
\begin{align}
\boldsymbol{A}^{-1} = \boldsymbol{L}\boldsymbol{L}^T,\qquad
\boldsymbol{x} = \boldsymbol{L} \boldsymbol{z},\qquad
z_j\sim \mathcal{N}(0, 1), \text{ for } j = 1,2,...,250.
\end{align}
%
where $\boldsymbol{L}$ is the Cholesky decomposition of the covariance matrix, $\boldsymbol{A}^{-1}$. Fig. \ref{fig:mvt}A shows the $R^*$ distribution resultant from 10,000 NUTS iterations in this case: now the distribution is centred on $R^*=1$ (mean $R^*=1.00$) indicating convergence. This was also reiterated by the fact that $\widehat{R}<1.01$ and $\text{ESS}>400$ for all parameters.

In GBMs, it is possible to calculate variable importance (see, for example, \cite{friedman2001greedy} and \cite{greenwell2019package}), and this allows us to determine which variables were mostly responsible for predictive power. We now compare these with the more established metrics $\widehat{R}$ and ESS as calculated in Stan. For a GBM fitted to the centered model, we plot in Fig. \ref{fig:mvt}B variable importance (here high values mean a variable is more important) versus $\widehat{R}$ for all dimensions of the target distribution (including Stan's $lp$ quantity, shown as a triangle). From this plot, there appears a weak positive association between GBM's variable importance and $\widehat{R}$, although the ranked correlation is not significant (Spearman's rank correlation: $\rho=0.05, S=2468800, p=0.35$). In Fig.  \ref{fig:mvt}C, we plot variable importance versus ESS, which exhibits a strong non-linear negative association (Spearman's rank correlation: $\rho=-0.54, S=4060003, p<0.01$). Since neither of these plots form perfect ``lines'' along which all the plotted points fall, this illustrates that variable importance provides information complementary to $\widehat{R}$ and ESS.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{../output/mvt_three.pdf}}
	\caption{\textbf{Multivariate normal example.} A shows $R^*$ distributions obtained for two MCMC samples from the centered parameterisation (``cp'') and one from the non-centered version (``ncp''); B shows variable importance versus $\widehat{R}$ for each parameter as calculated by Stan; and C shows variable importance versus ESS as calculated by Stan. In plots B and C, horizontal jitter was added to the points and a loess fit line with standard errors overlaid. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:mvt}
\end{figure}

\subsection{Wide datasets: multivariate normal}\label{sec:wide}
As the number of parameter dimensions increases, it could be hypthesised that ML algorithms will overfit the data, and, hence, testing set classification would be poor; in being so, this could lead us to falsely conclude that convergence had occurred. To test this hypothesis, we investigated two scenarios based around the multivariate normal. In the first of these, we used the 250-dimensional multivariate normal of eq. \ref{eq:mvt_normal_250} and used 250 post-warm-up iterations (after 250 warm-up iterations) from Stan to calculate $R^*$ distributions as in Algorithm \ref{alg:R_star_uncertainty}. Here, we considered both the centered and non-centered parameterisations, where, in both cases, the number of iterations is comparable to the number of parameters, so the training data is relatively ``wide'': the $R^*$ distribution in each case is shown in Fig. \ref{fig:mvt_wide_both}A. In this figure, it is clear that both $R^*$ distributions are shifted away from 1, indicating non-convergence. The same conclusion is reached if split-$\widehat{R}$ is used instead (Fig. \ref{fig:wide_both_diagnostics}A), since, for both parameterisations, there remain some parameters with $\widehat{R}>1.01$. Using bulk- or tail-ESS instead, we conclude that the non-centered parameterisation shows signs of convergence whereas the centered does not (Fig. \ref{fig:wide_both_diagnostics}B\&C).

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1\textwidth]{../output/wide_both_diagnostics.pdf}}
	\caption{\textbf{Established convergence diagnostics for 250-dimensional normal example.} The top row shows the results for the centered parameterisation; the bottom row for the non-centered. Column A shows split-$\widehat{R}$; columns B and C show the bulk- and tail-ESS; in each case the statistics are displayed for all model parameters and were calculated using 250 post-warm-up samples from Stan's NUTS algorithm. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:wide_both_diagnostics}
\end{figure}

We next consider a more challenging example -- a target distribution with 10,000 dimensions. In this case, we assume independent standard normals for each dimension. In Fig. \ref{fig:mvt_wide_both}B, we plot the $R^*$ distribution for two MCMC runs targeting this distributions: one with 400 iterations, the other with 1000. In both cases, the distributions were right of $R^*=1$, indicating non-convergence. These results were also echoed by split-$\widehat{R}$, with 28\% of dimensions having $\widehat{R}>1.01$ for the 400 iteration case and 3\% for the 1000 iteration case.
